---
title: "Additive Noise Models: Illustration and Hilbert Schmidt Independence Criterion (HSIC) Test"
author: "Anas Mourahib"
date: "`r Sys.Date()`"
output: html_document
---



```{r setup, include=FALSE}
if(!require(dHSIC)) install.packages("dHSIC", repos = "https://cloud.r-project.org/")
```

############################################################

## What is an Additive Noise Model?

An **Additive Noise Model (ANM)** assumes that the effect variable
can be written as

\[
Y = f(X) + N_Y
\]

where:

- \(X\) is the cause
- \(Y\) is the effect
- \(N_Y\) is a noise term
- \(N_Y\) is **independent** of \(X\)

A key idea of ANMs is that this independence typically holds
**only in the correct causal direction**.

In this document, we simulate data where the true causal direction
is:

\[
X \rightarrow Y
\]

and then compare it with the anti-causal direction.

---

############################################################
# DATA GENERATION
############################################################

```{r}
# ----------------------------------------------------------
# Step 1: Set random seed for reproducibility
# ----------------------------------------------------------
set.seed(123)

# ----------------------------------------------------------
# Step 2: Define parameters
# ----------------------------------------------------------
a <- 0      # lower bound of uniform distribution
b <- 1      # upper bound of uniform distribution
n <- 10^3   # sample size

# ----------------------------------------------------------
# Step 3: Generate cause X
# X is uniformly distributed on [0,1]
# ----------------------------------------------------------
X <- runif(n, a, b)

# ----------------------------------------------------------
# Step 4: Generate noise term N_Y
# Noise is independent of X by construction
# ----------------------------------------------------------
N_Y <- runif(n, a, b)

# ----------------------------------------------------------
# Step 5: Generate effect Y using an additive noise model
# True causal mechanism: Y = 0.5 * X + N_Y
# ----------------------------------------------------------
Y <- 0.5 * X + N_Y

```



# ----------------------------------------------------------

#  Causal direction X -> Y 

# ----------------------------------------------------------

Regression in the Correct Causal Direction

We now regress Y on X, which corresponds to the true causal
direction.

If the ANM assumption is correct, the regression residuals
should be independent of X.

############################################################

# Fit linear regression Y ~ X

############################################################
```{r}
causal_model <- lm(Y ~ X)

# ----------------------------------------------------------

# Step 7: Extract residuals

# These residuals estimate the noise N_Y

# ----------------------------------------------------------

residual_causal <- residuals(causal_model)




```

Plot: Residuals vs X

If residuals are independent of X, we should not see
any systematic pattern.

```{r}
plot(
X,
residual_causal,
xlab = "X (Cause)",
ylab = "Residuals of regression Y on X",
main = "Causal Direction: Y ~ X",
pch = 16,
col = rgb(0, 0, 1, 0.5)
)





```


############################################################

# Independence test- Hilbert--Schmidt Independence Criterion

############################################################




To test whether two samples $U$ and $V$ are independent, you can use the \emph{Hilbert--Schmidt Independence Criterion (HSIC)} test; Gretton, A., K. Fukumizu, C. H. Teo, L. Song, B. Sch\"olkopf and A. J. Smola (2007). A kernel statistical test of independence. In Advances in Neural Information Processing Systems (pp. 585-592).
This test considers the hypotheses
\[
H_0 : U \!\perp\!\!\!\perp V, \qquad
H_1 : U \not\!\perp\!\!\!\perp V.
\]

In \textsf{R}, the test can be performed using the \texttt{dHSIC} package as follows:


```{r}
library(dHSIC)

###Here, R uses automatically the permutation method and the Gaussian kernels for the HSIC test
dtes_causal = dhsic.test(X , residual_causal)
pvalue_causal = dtes_causal$p.value
cat(
  "This is the p-value for testing X independent of the residuals of Y on X:",
  pvalue_causal,
  ". Since this p-value is smaller than 0.05, we reject the null hypothesis H_0 of independence."
)



```

# ----------------------------------------------------------

# Anti-causal direction Y -> X 

# ----------------------------------------------------------

Regression in the Wrong Direction

Now we regress X on Y, corresponding to the anti-causal direction.

In general, the residuals will not be independent of Y.



```{r}

# ----------------------------------------------------------

# Step 8: Fit linear regression X ~ Y (anti-causal)

# ----------------------------------------------------------

anticausal_model <- lm(X ~ Y)

# ----------------------------------------------------------

# Step 9: Extract residuals

# ----------------------------------------------------------

residual_anticausal <- residuals(anticausal_model)



```

Plot: Residuals vs Y

Dependence patterns typically appear here.


```{r}
plot(
Y,
residual_anticausal,
xlab = "Y (Effect)",
ylab = "Residuals of regression X on Y",
main = "Anti-Causal Direction: X ~ Y",
pch = 16,
col = rgb(1, 0, 0, 0.5)
)





```


############################################################

# Independence test: Hilbert--Schmidt Independence Criterion

############################################################



```{r}
library(dHSIC)


dtes_anticausal = dhsic.test(Y , residual_anticausal)
pvalue_anticausal =dtes_anticausal$p.value
cat(
  "This is the p-value for testing Y independent of the residuals of X on Y:",
  pvalue_anticausal,
  ". Since this p-value is smaller than 0.05, we reject the null hypothesis H_0 of independence."
)



```


############################################################

# Results

############################################################

The ANM method validates the true causal model X -> Y. Note that this is a simple case where the function $f$ is linear. Another challenging case would be if $f$ is not linear which you can find in another file 